Multimodal spatial availability: a singly-constrained measure of accessibility considering multiple modes

### 2023-11-07 Journal-manuscript/Journal-manuscript.Rmd
Further edits, close to 80% done

The paper rest of this paper is organized as follows. In Section we use a synthetic examples with multiple modes to illustrate some relevant issues; this helps us motivate the derivation of the new spatial availability expressions. This is followed in Section 3 by an empirical example using data from the city of Madrid after the implementation of its Low Emission Zones. Data for this example comes from the city's 2018 travel survey. The example shows the differences in spatial availability within and outside the LEZ for travellers using different modes, namely car, transit, cycling and walking. In Section 4, we provide concluding remarks on the strengths of the use of spatial availability as a multimodal accessibility measure, and discuss potential future uses in policy planning scenarios as well as directions for future research.

@@

# A review of multimodal accessibility measures

Location-based accessibility indicators are quantitative measures of _potential_ interaction with opportunities for locations within a given region: they are summary measures of the relationship between land-use and transport systems. Arguably, the most commonly used are measures based on the gravity model [@wuUnifyingAccess2020], of which cumulative opportunity measures and weighted cumulative opportunity measures are particular forms [@levinsonTransportAccessManual2020]. These measures assign a weight to opportunities based on how easy it is to reach them. Given an origin ($i$) and a destination ($j$), an impedance function $f^{m}(c^m_{ij})$ converts the cost of travel (e.g., time, money, generalized cost) into a score that represents the propensity for interaction. These measures originate from that proposed by @hansenHowAccessibilityShapes1959, which can take the following form in the multimodal case: $S_i^m = \sum_j O_j f^m(c_{ij}^m)$ where $m$ is a set of modes which have mode-specific travel costs ($(c_{ij}^m)$) and/or travel impedance functions ($f^m(\cdot)$).

The Hansen-type measure is not constrained, and as a result does not consider the opportunities as finite. As an example, the work of @tahmasbiMultimodalAccessibilitybasedEquity2019 uses the Hansen-type measure to measure the potential interaction with retail locations using walking, public transit, and car modes as $m$. $S_i^m$ is the sum of retail locations $j$ that can potentially be reached under the travel impedance as calculated for each $i$ and $m$. In other words, each $i$ has three $S_i^m$ values, one per $m$. In this work, they demonstrate that the car mode has the highest $S_i^{m}$ values in the majority of $i$, i.e., populations using a car can potentially reach more retail opportunities than populations using other modes. However, higher $S_i^{m}$ values for car do not affect the values of $S_i^{m}$ for other modes: in effect, each modes is analyzed as if the others did not exist, which would be reasonable for an inexhaustible opportunity, but is more questionable when the opportunity is subject to congestion. Since the measure is measure is also not constrained, and each opportunity can be and often is counted multiple times within and between modes, the sum of accessibility is not a meaningful quantity. To improve interpretability, $S_i^m$ values are presented as standardized accessibility index in this work, from the lowest value of 0 to the highest value of 1. This presentation or discussion of Hansen-type accessibility measures is common in the literature [e.g., in @campbell2019accessibility; and @maharjanSpatialEquityModal2022].

However, this representation of results can make the interpretation of results challenging _because_ values are region-relative and lacking numeric meaning. Zones will always have values from between 0 to 1 represented, but is a zone with a low value for pedestrian-modes and a high value for car-modes notable? And if notable, what does the difference in these standardized values tangibly mean for planners? By how much should transport systems and land-use configurations be changed to improve conditions? And in what way can these existing methods be used to track progress over time? Or between regions? This is challenging to track, since certain values will always be relatively 'low' and 'high'. Alternatively, modal accessibility calculated using the Hansen-type measure is also presented without standardization, as in the work by @lunkeModalAccessibilityDisparities2022. Often, the meaning of the accessibility value itself is not interpreted, as it is difficult to state what having access to 30,000 employment opportunities by car but having 10,000 employment opportunity by transit tangibly signifies.

However, once opportunities in a region are interpreted as finite, a new meaning is taken on. As considered in the long tradition of accessibility work, capacity of opportunities is limited and thus is subject to competition by population [@soukhovIntroducingSpatialAvailability2023; @grengsJobAccessibilityModal2010; @kawabataJobAccessibilityIndicator2006; @kwokUseModalAccessibility2004; @morrisAccessibilityIndicatorsTransport1979; @weibullAxiomaticApproachMeasurement1976; @shenLocationCharacteristicsInnercity1998]. There are only so many school-seats, hospital capacity, employment opportunities, etc., in a region and if one person interacts with an opportunity at a given time, it is taken: the supply of an opportunity and the demand for that opportunity are two components of accessibility. The competition consideration is especially important in the case of multiple modes. For instance, people in a zone who are advantaged with a relatively low travel-cost mode have the ability to potentially interact with more opportunities than other people. Due to this advantage, through the perspective of finite opportunities, there are fewer opportunities left to be potentially interacted for everyone else, especially zones with populations using higher travel-cost modes. This recognition is the motivation behind integrating _competition_ for opportunities within modal accessibility measures.

Arguably, one of the most popular competitive location-based accessibility measures is the two-step floating catchment area (2SFCA) approach popularized by @luoMeasuresSpatialAccessibility2003 who simplified the approach proposed by @shenLocationCharacteristicsInnercity1998 (with similar considerations for competition in @weibullAxiomaticApproachMeasurement1976 and @josephMeasuringPotentialPhysical1982). The Shen-type accessibility measure's formulation is: $a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}$ where $D_j^m$ is the potential demand for opportunities equal to travel impedance weighted population $\sum_i P_i^m f^m(c_{ij}^m)$ and the remaining variables are repeated in the Hansen-type measure. The Shen-type measure per mode ($a_i^m$) can be understood as a ratio of the travel impedance-weighted supply of opportunities for the $m$-using population in $i$ over the ravel impedance-weighted demand for opportunities. In this way, it considers competition but the measure remains _non-constrained_ meaning resulting values are not associated with global maximums. In other words, it is difficult to interpret the meaning of differences in Shen-type accessibility scores between modes in similar ways as reflected for Hansen-type measures.

To illustrate, @taoInvestigatingImpactsPublic2020a calculates $a_i^m$ to jobs for different income-group populations in Shenzhen, China using $m = \text{public transit}$ and $m={car}$. They demonstrate that $i$ with low-income populations have lower $a_i^m$ than $i$ with higher-income populations. Further, they demonstrate that $a_i^{m=\text{public transit}}$ is lower than $a_i^{m=\text{car}}$ at many $i$, arguing that this may put $i$ with lower-income populations in a further disadvantage. $a_i$ and/or $a_i^m$ are used to compare relative spatial differences in overall competitive accessibility and multimodal competitive accessibility, but because there is no global maximum, making it is difficult to interpret the significance between differences in $a_i^{m}$ values. Questions such as: what is the impact that competition has on the difference in $a_i^m$ values? How does impact vary spatially? And what is the interpretation of this difference? are left unaddressed.

Spatial availability improves on previously discussed multimodal accessibility approaches using the Hansen-type measure and the Shen-type measure by taking into account _competition_ in the potential interaction with opportunities in a _constrained_ framework (e.g., finite opportunities). This is done by considering: 1) competition between mass effect (e.g., the advantage of sub-populations residing in relatively low population-density and high opportunity-proximate areas) and 2) competition between travel impedance (e.g., sub-populations with relatively low travel-impedance) through a proportional allocation mechanism. The following sub-section demonstrates how spatial availability compares to the Hansen-type and Shen-type measures through a synthetic example.

## Multimodal spatial availability $V_i^m$

In brief, we define the spatial availability at $i$ ($V_{i}$) as the proportion of all opportunities in the region $O$ that are allocated to location $i$ from all opportunity destinations $j$. $V_{i}$ is a value of how many opportunities are available to each location $i$ out of all the opportunities in the region. The general formulation of spatial availability $V_{i}$ is shown in Equation (\ref{eq:spatial-availability-general}):
```{=tex}
\begin{equation}
\label{eq:spatial-availability-general}
V_i = \sum_{j=1}^J O_jF^t_{ij}
\end{equation}
```

@@

- $F^t_{ij}$ is a balancing factor that depends on the demand for opportunities $O_j$ and cost of movement in the system $f(c_{ij})$.
- $V_i$ is the number of spatially available opportunities at $i$; the sum of $V_{i}$ is equivalent to the total sum of opportunities in the region (i.e., $\sum_j O_j = \sum_i V_i$)

The spatial availability measure is introduced in @soukhovIntroducingSpatialAvailability2023. Spatial availability's unique feature is the balancing factor $F^t_{ij}$, a proportional allocation mechanisms, that ensures the $V_i$ calculated for each $i$ sums, across all $i$ in the region, to equal the total number of opportunities in the region. As such, spatial availability is a  _competitive_ and _constrained_ accessibility measure as $F^t_{ij}$ handles the number of opportunities in the region in a finite way (proportional allocation). $F^t_{ij}$ consists of two components: a population-based balancing factor $F^p_{i} = \frac{P_i}{\sum_i P_i}$ and an impedance-based balancing factor $F^c_{ij} = \frac{F^c_{ij}}{\sum_j F^c_{ij}}$ that, respectively, allocates opportunities to $i$ in proportion to the size of the population at $i$ (the mass effect) and the cost of reaching opportunities at $j$ (the impedance effect).

$F^p_{i}$ and $ F^c_{ij}$ are calculated for each $i$ such that they both equal 1 when summed across all $i$ in the region (e.g., $\sum_i F^p_{i} = 1$ and $\sum_i F^c_{ij} = 1$). These balancing factors are combined multiplicatively to yield $F^t_{ij}$ which ensures that a proportion of the opportunities $O_j$ are allocated to each $i$ accordingly. In other words, assuming a finite number of opportunities in the region, $F^t_{ij}$ proportionally allocates $O_j$ to each $i$ such that the resulting $V_i$ value represents the number of opportunities _spatially available_ to the population at $i$. Each zonal value is a proportion of the opportunities in the region (i.e., $\sum_j O_j = \sum_i V_i$).

The focus of this paper is to extend $V_i$ for the measurement of multimodal accessibility applications. To do so, the balancing factors are reformulated to yield a proportional value for the set of modes $m$ used by populations at each $i$. As these factors are proportional, $F^{pm}_{i}$ and $F^{cm}_{ij}$ can be summed across each $m$ at each $i$ and then across all $i$ to equal to 1. They are also similarly combined multiplicatively to obtain their joint effect, represented as the combined balancing factor $F^{tm}_{ij}$ detailed in Equation (\ref{eq:multimodal-balancing-factors}).

```{=tex}
\begin{equation}
\label{eq:multimodal-balancing-factors}
F^{tm}_{ij} = \frac{F^{pm}_{i} \cdot F^{cm}_{ij}}{\sum_{m=1}^M \sum_{i=1}^N F^{pm}_{i} \cdot F^{cm}_{ij}}
\end{equation}
```

@@

- The population balancing factor for each $m$ at each $i$ is $F^{pm}_{i} = \frac{P_{i}^m}{\sum_{m}\sum_{i} P_{i}^m}$
- The cost of travel balancing factor for each $m$ at $i$ is $F_{ij}^{cm} = \frac{f^m(c_{ij}^m)}{\sum_{m} \sum_{i} f^m(c_{ij}^m)}$

Implementing $F^{tm}_{ij}$, the following Equation (\ref{eq:spatial-availability-multimodal}) demonstrates the multimodal configuration of spatial availability $V_i^m$:

```{=tex}
\begin{equation}
\label{eq:spatial-availability-multimodal}
V^m_{i} = \sum_{j=1}^J O_j\ F^{tm}_{ij}
\end{equation}
```

-   $m$ is a set of modes used by populations in the region.
-   $F^{tm}_{ij}$ is a balancing factor $F^t_{ij}$ for each $m$ at each $i$.
-   $V^m_{i}$ is the spatial availability $V_{i}$ for mode $m$ at each $i$; the sum of $V^m_{i}$ for all $m$ at each $i$ is equivalent to the total sum of opportunities in the region (i.e., $\sum_j O_j = \sum_i V_i = \sum_{m} \sum_{i} V^m_{i}$)

## An illustrative synthetic example
Consider the following Figure \ref{fig:Fig1}, it depicts a region with population and jobs at three population centers ($A$, $B$, $C$) and three employment centers ($1$, $2$, $3$). The population at each population center is divided into two sub-populations, one using a faster mode $z$ and another using a slower mode $x$, to travel to employment centers. Population center $A$ is Suburban: it is closest to its own relatively large employment center at $1$, close to the Urban's equally large employment center $2$, and has a population that is smaller than the Urban $B$ and larger than the Satellite $C$. $B$ has the largest $x$-using population, followed by then $A$, then $C$. This synthetic example was inspired by the single-mode example used in @shenLocationCharacteristicsInnercity1998 and reconfigured in @soukhovIntroducingSpatialAvailability2023.

@@

From the perspective of access to a _finite_ amount of opportunities in the region ($210,000$ jobs), the sub-population that is most proximate to jobs, furthest from densely populated centers, and is using the lowest travel-cost mode $z$ can potentially access the most job opportunities. This appears to be the sub-population at $A$ using mode $z$. Sub-populations located in opposite conditions (i.e., further away from jobs, close to dense populations, and using $x$) are at a relative job opportunity access _disadvantage_. The competition for opportunities between different mode-using populations matters from the perspective of inequities as it reflects how well the land-use and transport system serves (or doesn't serve) certain populations.

@@

The calculated $S_i^m$, $a_i^m$ and $V_i^m$ accessibility values for each $i$ and $m$ are shown in the middle three columns and are aggregated for each $i$ in the final two columns in Table 1 <!-- \ref{tab:tab1}-->. We use a negative exponential impedance function $f^m(c_{ij}^m) = \exp(-\beta\cdot c_{ij})$ with $\beta=0.1$ for both $x$ and $z$ modes for all accessibility measures calculations.

The Hansen-type measure $S_i^m$ is presented for each origin and mode in third column of Table 1 <!-- \ref{tab:tab1}-->. For all $i$, the $z$-using sub-population has higher $S_i^m$ values than the $x$-using sub-populations. Additionally, $S_i^m$ is equal for both mode-using populations in $A$ and $B$. This is the case since $S_i^m$ does not consider _competition_, it only relies on reflecting the count of opportunities that may be interacted with as a product of $f^m(c_{ij}^m)$. Recall, populations in $A$ and $B$ have the same travel impedance to employment centers $1$, $2$ and $3$ (either 15, 30, or 100 minutes using $x$ or 10, 25, or 80 minutes using $z$). As such, these the calculated $S_i^m$ values are the same for both $A$ and $B$. Furthermore, the total sum of $S_i^m$ in the region is equal to `r S_im$S_im %>% sum() %>% prettyNum(big.mark = ",")`. This value is difficult to interpret: it represents the weighted sum of opportunities that may be interacted with within the region based on travel impedance. It cannot be interpreted as any sort of benchmark since the measure is _non-constrained_. To connect this example to literature, $S_i^m$ is calculated in the work of @tahmasbiMultimodalAccessibilitybasedEquity2019; they compare differences in $S_i^m$ values between modes in a relative and comparative sense, but make no further interpretation of the $S_i^m$ values.

In the fourth and sixth column in Table 1 <!-- \ref{tab:tab1}--> the Shen-type measure is calculated: first for both origin and mode $a_i^m$ as well as aggregated by the weighted mean mode-population ( $\sum_m \frac{P_i^m}{P_i}*a_i^m$ ) to represent a value for each origin $a_i$. Unlike $S_i^m$, this measure considers _competition_. For instance, the $x$-using populations in $A$ and $B$ centers do not have the same $a_i^m$ values as the $z$-using. In fact, $A$ has the highest values $a_i^m$ and $a_i$ values since this center has the smallest travel impedance to opportunities (lower than at $C$, $A$ and $B$ are equal) and has one of these lowest proximity to a relatively high amount of population (lower than at $B$).

>>>>

The paper rest of this paper is organized as follows. In Section we discuss the use a synthetic examples with multiple modes to illustrate some relevant issues; this helps us motivate the derivation of the new spatial availability expressions. This is followed in Section 3 by an empirical example using data from the city of Madrid after the implementation of its Low Emission Zones. Data for this example comes from the city's 2018 travel survey. The example shows the differences in spatial availability within and outside the LEZ for travelers using different modes, namely car, transit, cycling and walking. In Section 4, we provide concluding remarks on the strengths of the use of spatial availability as a multimodal accessibility measure, and discuss potential future uses in policy planning scenarios as well as directions for future research.

@@

# A brief review of multimodal accessibility

Location-based accessibility indicators are quantitative measures of _potential_ interaction with opportunities for locations within a given region: they are summary measures of the relationship between land-use and transport systems. Arguably, the most commonly used are measures based on the gravity model [@wuUnifyingAccess2020], of which cumulative opportunity measures and weighted cumulative opportunity measures are particular forms [@levinsonTransportAccessManual2020]. These measures assign a weight to opportunities based on how easy it is to reach them. Given an origin ($i$) and a destination ($j$), an impedance function $f^{m}(c^m_{ij})$ converts the cost of travel (e.g., time, money, generalized cost) into a score that represents the propensity for interaction. These measures originate from that proposed by @hansenHowAccessibilityShapes1959, which can take the following form in the multimodal case: $S_i^m = \sum_j O_j f^m(c_{ij}^m)$ where $m$ is a set of modes which have mode-specific travel costs ($c_{ij}^m$) and/or travel impedance functions ($f^m(\cdot)$).

Hansen-type accessibility is not constrained, which is to say it does not consider the opportunities as finite. To cite an example, Tahmasbi et al. [@tahmasbiMultimodalAccessibilitybasedEquity2019] use Hansen-type accessibility to assess the potential interaction with retail locations by three modes: walking, public transit, and car (i.e., $m = w, p, c$). $S_i^m$ is the sum of retail locations $j$ that can potentially be reached under the travel impedance as calculated for each $i$ and $m$. In other words, for each origin $i$ three accessibility scores are calculated. In this work, Tahmasbi et al. [@tahmasbiMultimodalAccessibilitybasedEquity2019] show that car travel affords the highest $S_i^{m}$ values in the majority of $i$, i.e., travelers who use a car can potentially reach more retail opportunities than populations using other modes. However, higher $S_i^{m}$ values for car do not affect the values of $S_i^{m}$ for other modes: in effect, each mode is analyzed as if the others did not exist. Since the measure is not constrained, each opportunity is typically counted multiple times within and between modes, and as a result the sum of accessibility is not necessarily a meaningful quantity. The accessibility scores for the modes are often values that are difficult to interpret beyond making statements about relative size. For example, Lunke [@lunkeModalAccessibilityDisparities2022], researching the region of Oslo, reports accessibility scores for car in the order of tens of thousands of employment opportunities. The corresponding scores for transit are lower, but still often in the thousands or tens of thousands. As reported, the ratio of the transit to the car score can be lower than 0.2 (meaning transit gives access to less than 20% of the opportunities than car). But despite the discussion about "sufficient accessibility" (e.g., in the abstract), it is unclear what the unconstrained scores mean: is having access to but 10,000 jobs by transit insufficient? After all, 10,000 employment opportunities are still plenty of opportunities. These ratios can be found elsewhere in the literature [e.g., @paezRelative2010, Figs. 7, 8, and 9; @paezHealth2010, Fig. 5; @paezJobs2013, Figs. 6, 7, and 8], and they are useful to assess when some members of the public are better or worse off than others, but they do not say much about how bad is "worse".

Besides ratios of accessibility, another way to improve interpretability of scores sometimes seen in the literature is to standardize them to lie in the range [0-1]. This adjustment is only helpful insofar as it facilitates relative comparisons, but interpretation of the scores remains challenging because the values are specific to a region and convey no meaning about the magnitude of the scores. In this approach, zones always have values between 0 and 1, but how remarkable is a zone with a low score for pedestrians and a high value for car? And if remarkable, what does the difference in these standardized values mean for planners? By how much should transport systems and land-use configurations be changed to improve conditions? And in what way can these scores be used to track differences over time? Or between regions? These questions lack straightforward answers since certain values will always be relatively 'low' or 'high', but do not track to a quantity that can be intuitively understood. Presentation or discussion of Hansen-type accessibility that has been standardized in this way is not uncommon in the literature [e.g., @campbell2019accessibility; @maharjanSpatialEquityModal2022].

Once we understand opportunities to be finite, it is possible for an accessibility measure to take a crisper meaning. As considered in the long tradition of accessibility research, capacity of opportunities is limited and thus is subject to competition by population [@soukhovIntroducingSpatialAvailability2023; @grengsJobAccessibilityModal2010; @kawabataJobAccessibilityIndicator2006; @kwokUseModalAccessibility2004; @morrisAccessibilityIndicatorsTransport1979; @weibullAxiomaticApproachMeasurement1976; @shenLocationCharacteristicsInnercity1998]. There are only so many school-seats, hospital capacity, employment opportunities, etc., in a region and if one person reaches an opportunity, it is taken: the supply of an opportunity and the demand for that opportunity are two components of accessibility. These are clear examples of opportunities that are clearly competitive. But we would go as far as to argue that every type of opportunity is subject to congestion or capacity constraint, even when the opportunities are conventionally seen as non-competitive. Amenities are a good example of this. For instance, standards for providing green spaces are often stated in the form of _exclusive access_, in units of amenity per capita. A case in point is Natural England, an organization that recommends an Accessible Natural Greenspace Standard such that the minimum supply of space is one ha of statutory Local Nature Reserves per thousand population^[see  https://redfrogforum.org/wp-content/uploads/2019/11/67-Nature-Nearby%E2%80%99-Accessible-Natural-Greenspace-Guidance.pdf]. Similarly, the World Health Organization [cited in @OECDFrameworks2013] recommends that cities provide a minimum of 9$m^2$ of green area per inhabitant. For our purposes, standards of this type translate into "how much of a resource can be offered to each individual that is not allocated to anyone else?". Green spaces often have large capacities, but they still have a capacity. Constraining accessibility is in this way a useful way to evaluate the congested availability of any type of opportunity. As development of sound standards is emphasized in the planning literature, in particular in regards to fairness in transportation [see @martensFair2021], spatial availability analysis is a useful way to develop and assess standards.

The relevance of the considerations above is put in sharper relief when we think about the use of multiple modes (or heterogeneous populations). If we return to Oslo for a moment [@lunkeModalAccessibilityDisparities2022], we notice that the places that have high accessibility by transit are also the places that have _very high_ accessibility by car (Figure 2). Those two populations are going for the same opportunities, and those travelling by transit have fewer to choose from to begin with. More generally, people in a zone who are advantaged with relatively cost of travel will have the ability to potentially reach more opportunities than other people. Due to this advantage, through the perspective of finite opportunities, there are fewer opportunities left for everyone else, especially for those who use modes that are slower or more expensive.

As noted in the introduction, competitive accessibility was the rationale for developing floating catchment area methods (FCA), popularized by @luoMeasuresSpatialAccessibility2003 [although similar, and earlier, developments are found in @weibullAxiomaticApproachMeasurement1976; @josephMeasuringPotentialPhysical1982; @shenLocationCharacteristicsInnercity1998]. Shen-type accessibility is formulated as: $a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}$ where $D_j^m$ is the potential demand for opportunities equal to travel impedance weighted population $\sum_i P_i^m f^m(c_{ij}^m)$ and the remaining variables are repeated in the Hansen-type measure. Shen-type modal accessibility ($a_i^m$) can be understood as a ratio of the travel impedance-weighted supply of opportunities for $m$-mode in $i$ over the travel impedance-weighted demand for opportunities. In this way, it considers competition. That said, the measure remains unconstrained, meaning both population _and_ opportunities are multiply counted [see @paezDemand2019]. In other words, interpretation of differences in Shen-type accessibility scores between modes is fraught as it is for Hansen-type measures.

To illustrate, @taoInvestigatingImpactsPublic2020a calculate $a_i^m$ to jobs for different income-group populations in Shenzhen, China using $m = \text{public transit}$ and $m=\text{car}$. Their results indicate that zones with low-income populations have lower $a_i^m$ than zones with higher-income populations. Further, they show that $a_i^{m=\text{public transit}}$ is lower than $a_i^{m=\text{car}}$ at many zones, arguing that this may further place those zones with lower-income populations at disadvantage. $a_i$ and/or $a_i^m$ are used to compare relative spatial differences in overall competitive accessibility and multimodal competitive accessibility, but because opportunities were doubly counted (entering the sums of both modes), this makes for uneasy interpretations of the differences in $a_i^{m}$ between modes. Questions that this approach leaves unaddressed include: what is the impact of competition on the difference in $a_i^m$ values? How does the impact vary spatially? And what is the interpretation of this difference?

Spatial availability improves on previously discussed accessibility approaches using the Hansen-type measure and the Shen-type measure by constraining the sum of opportunities, that is, by treating opportunities as finite. This is done by means of proportional allocation factors that follow well established principles of spatial interaction and the gravity model [see @wilsonFamilty1971]. In Soukhov et al. [@soukhovIntroducingSpatialAvailability2023] these factors consider: 1) the mass effect (e.g., the size of populations at different origins); and 2) the cost of travel from different zones (e.g., some sub-populations face relatively higher or lower costs). The following section introduces the multimodal form of spatial availability.

# Multimodal spatial availability $V_i^m$

In brief, we define the spatial availability at $i$ ($V_{i}$) as the proportion of all opportunities in the region $O$ that are allocated to location $i$ from all opportunity destinations $j$. $V_{i}$ is a value of how many opportunities are available to each location $i$ out of all the opportunities in the region. The general formulation of spatial availability $V_{i}$ is shown in Equation (\ref{eq:spatial-availability-general}) [see @soukhovIntroducingSpatialAvailability2023]:
```{=tex}
\begin{equation}
\label{eq:spatial-availability-general}
V_i = \sum_{j=1}^J O_jF^t_{ij}
\end{equation}
```

@@

- $F^t_{ij}$ is a balancing factor that depends on the size of the populations at different locations that demand opportunities $O_j$, as well as the cost of movement in the system $f(c_{ij})$.

- $V_i$ is the number of spatially available opportunities at $i$; the sum of $V_{i}$ is identical to the total number of opportunities in the region (i.e., $\sum_j O_j = \sum_i V_i$); in other words, opportunities are dealt with as finite resources.

Compared to Hansen-type accessibility:
```{=tex}
\begin{equation}
\label{eq:hansen-accessibility}
A_i = \sum_{j=1}^J O_jf(c_{ij})
\end{equation}
```

\noindent we see that spatial availability is, like Hansen's measure, a weighted sum of the opportunities.
What makes spatial availability stand apart from other approaches is the weight used in the sum, balancing factor $F^t_{ij}$, implements a proportional allocation mechanisms to ensure that the sum of $V_i$ is constrained to match the total number of opportunities in the region. As such, spatial availability is singly-constrained and naturally implements competition or congestion. $F^t_{ij}$ consists of two parts. The first part is a population-based proportional allocation factor to model the mass effect of the gravity model:
$$
F^p_{i} = \frac{P_i}{\sum_i P_i}
$$

This factor makes opportunities available based on demand. Secondly, there is an impedance-based proportional allocation factor that model the cost effect:
$$
F^c_{ij} = \frac{F^c_{ij}}{\sum_j F^c_{ij}}
$$

This factor makes opportunities available preferentially to those who can reach them at a lower cost. $F^p_{i}$ and $F^c_{ij}$ are designed so that they both equal 1 when summed across all $i$ in the region (e.g., $\sum_i F^p_{i} = 1$ and $\sum_i F^c_{ij} = 1$). These factors are combined multiplicatively to yield $F^t_{ij}$ which ensures that a proportion of the opportunities $O_j$ are allocated to each $i$ accordingly. In other words, assuming a finite number of opportunities in the region, $F^t_{ij}$ proportionally allocates $O_j$ to each $i$ such that the resulting $V_i$ value represents the number of opportunities _available_ to the population at $i$. Each zonal value is a proportion of the opportunities in the region (i.e., $\sum_j O_j = \sum_i V_i$).

The focus of this paper is to extend $V_i$ for the measurement of multimodal applications (or more generally heterogeneous populations). To do so, the balancing factors need to be reformulated so that 1) the mass effect now accounts not only for the size of the population at $i$, but also the size of sub-populations within $i$; and 2) the cost of travel is not only for different zones, but by sub-populations within each zone (e.g., the cost of travel from $i$ by car, transit, walking, etc.) When we introduce modes (or sub-populations) $m$, the proportional allocation factors need to satisfy the condition that $F^{pm}_{i}$ and $F^{cm}_{ij}$ can be summed across each $m$ at each $i$ and then across all $i$ to equal to 1. They are also similarly combined multiplicatively to obtain their joint effect, represented as the combined balancing factor $F^{tm}_{ij}$ similar to that detailed in Equation (\ref{eq:multimodal-balancing-factors}). This factor is given by:
```{=tex}
\begin{equation}
\label{eq:multimodal-balancing-factors}
F^{tm}_{ij} = \frac{F^{pm}_{i} \cdot F^{cm}_{ij}}{\sum_{m=1}^M \sum_{i=1}^N F^{pm}_{i} \cdot F^{cm}_{ij}}
\end{equation}
```

\noindent where:

- The factor for allocation by population for each $m$ at each $i$ is $F^{pm}_{i} = \frac{P_{i}^m}{\sum_{m}\sum_{i} P_{i}^m}$; and
- The factor for allocation by cost of travel for each $m$ at $i$ is $F_{ij}^{cm} = \frac{f^m(c_{ij}^m)}{\sum_{m} \sum_{i} f^m(c_{ij}^m)}$

Implementing $F^{tm}_{ij}$, the following Equation (\ref{eq:spatial-availability-multimodal}) gives the multimodal version of spatial availability $V_i^m$:
```{=tex}
\begin{equation}
\label{eq:spatial-availability-multimodal}
V^m_{i} = \sum_{j=1}^J O_j\ F^{tm}_{ij}
\end{equation}
```

@@

-   $m = 1, 2,\cdots, M$ is a set of $M$ modes (or sub-populations) of interest.
-   $F^{tm}_{ij}$ is a balancing factor $F^t_{ij}$ for each $m$ at each $i$.
-   $V^m_{i}$ is the spatial availability $V_{i}$ for mode $m$ at each $i$; the sum of $V^m_{i}$ for all $m$ at each $i$ is equivalent to the total sum of opportunities in the region (i.e., $\sum_j O_j = \sum_i V_i = \sum_{m} \sum_{i} V^m_{i}$).

Next we use a synthetic example to contrast multimodal accessibility and spatial availability.

# An illustrative synthetic example

Consider the simple system shown in Figure \ref{fig:Fig1}. The figure shows a region with population at three population centers ($A$, $B$, $C$) and jobs at three employment centers ($1$, $2$, $3$). The population at each origin $i$ is consists of two sub-populations, one using a faster mode $z$ and another using a slower mode $x$, to travel to employment centers. Population center $A$ is Suburban: it is closest to its own relatively large employment center at $1$, close to the Urban's equally large employment center $2$, and has a population that is smaller than the Urban $B$ and larger than the Satellite $C$. $B$ has the largest $x$-using population, followed by then $A$, then $C$. This synthetic example is inspired by the single-mode example used in @shenLocationCharacteristicsInnercity1998 and reconfigured in @soukhovIntroducingSpatialAvailability2023.

@@

From the perspective of access to a _finite_ amount of opportunities in the region ($210,000$ jobs), the sub-population that is most proximate to jobs (low cost to reach), furthest from large populations (less competition), and uses the fastest mode $z$ (greater range) can potentially reach the largest number of opportunities. This appears to be the sub-population at $A$ using mode $z$. Sub-populations located in opposite conditions (i.e., more distant from jobs, close to large populations, and using slow mode $x$) are at a relative disadvantage. The competition for opportunities between different mode-using populations matters as it reflects how well the land-use and transport system serves (or does not serve) certain populations.

@@

The values calculated for $S_i^m$ (Hansen-type accessibility), $a_i^m$ (Shen-type accessibility), and $V_i^m$ (spatial availability) for each $i$ and $m$ are shown in the middle three columns and are aggregated for each $i$ in the final two columns in Table 1 <!-- \ref{tab:tab1}-->. As in the example in Shen [@shenLocationCharacteristicsInnercity1998], we use a negative exponential impedance function $f^m(c_{ij}^m) = \exp(-\beta\cdot c_{ij})$ with $\beta=0.1$ for both $x$ and $z$ modes for all accessibility measures calculations. Notice that in this example we use the same impedance function but the travel times are different for the two modes. More generally, it is possible to use different impedance functions for the modes, as demonstrated in the empirical example below.

Hansen-type accessibility $S_i^m$ is presented for each origin and mode in the third column of Table 1 <!-- \ref{tab:tab1}-->. For all $i$, the travel by $z$ results in higher values of $S_i^m$ than travel by $x$. Lack of competition, or alternatively the assumption of an inexhaustible resource in the calculation of $S_i^m$, lead to a curious result. Since the populations in $A$ and $B$ have the same travel impedance to employment centers $1$, $2$ and $3$ (either 15, 30, or 100 minutes using $x$ or 10, 25, or 80 minutes using $z$), their values of $S_i^m$ are the same for both $A$ and $B$. Furthermore, the total sum of $S_i^m$ in the region is equal to `r S_im$S_im %>% sum() %>% prettyNum(big.mark = ",")`. This value lacks an intuitive interpretation: it represents the weighted sum of opportunities that may be reached within the region according to the travel impedance (i.e., the travel behavior and the characteristics of the modes) and does not usefully translate into any sort of benchmark. To connect this example to the literature, $S_i^m$ is calculated in the work of @tahmasbiMultimodalAccessibilitybasedEquity2019; they compare differences in $S_i^m$ values between modes in a relative and comparative sense, but make no further interpretation of the $S_i^m$ values. Large metropolitan regions will have large values and small regions small, but that is simply an artifact of size.

In the fourth and sixth columns in Table 1 <!-- \ref{tab:tab1}--> the results for Shen-type accessibility are reported: first for both origin and mode $a_i^m$ as well as aggregated by the weighted mean mode-population ( $\sum_m \frac{P_i^m}{P_i}*a_i^m$ ) to represent a value for each origin $a_i$. Unlike $S_i^m$, this measure does considers _competition_. For instance, the population travelling by $x$ from $A$ and $B$ do not have the same values of $a_i^m$ as those travelling by $z$. In fact, $A$ has the highest values $a_i^m$ and $a_i$ values since this center has the lowest travel impedance to opportunities (lower than at $C$, $A$ and $B$ are equal) and has faces relatively low competition, not being close to a relatively large population (lower than at $B$).

However, the calculations of $a_i^m$ are not constrained: the total sum of $a_i^m$ or $a_i$ is practically meaningless since it represents a sum of ratios. For instance, the population travelling by $z$ from $A$ has a value of `r a_i %>% filter(i=="A" & m=="z") %>% select(a_im) %>% pull() %>% round(digits=2)` jobs per job-seeking population compared to `r a_i %>% filter(i=="A" & m=="x") %>% select(a_im) %>% pull() %>% round(digits=2)` for users of mode $x$. What is the meaning of these values? The difference between these modes is equal to `r ((a_i %>% filter(i=="A" & m=="z") %>% select(a_im)) - (a_i %>% filter(i=="A" & m=="x") %>% select(a_im))) %>% pull() %>% round(digits=2)`, but `r ((a_i %>% filter(i=="A" & m=="z") %>% select(a_im)) - (a_i %>% filter(i=="A" & m=="x") %>% select(a_im))) %>% pull() %>% round(digits=2)` of what? How many more job opportunities can users of $z$ reach compared to user of $x$? When $a_i^m$ is aggregated to $a_i$ as shown in the sixth column, the values face similar interpretability issues. 
The Shen-type measure is implemented in aforementioned work of Tao et al. [@taoInvestigatingImpactsPublic2020a] to calculate modal $a_i^m$ values and the aggregated $a_i$ is implemented in the work of [@carpentieriMultimodalAccessibilityPrimary2020]. However, similar to Hansen-type accessibility, these works discuss relative and spatially comparative differences in values, but veer from interpretating the values of $a_i^m$ or $a_i$ themselves. In fairness, interpretation is complicated by the multiple counting of opportunities between zones and modes.
